---
title: "Creating Custom Road Datasets with trafficCAR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating Custom Road Datasets with trafficCAR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---  
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette is written for **users of `trafficCAR`** who want to work with
road networks beyond the bundled examples.

You will learn how to:

1. Obtain road network data for a city
2. Save it as a GeoJSON file
3. Convert the GeoJSON into an `.rda` dataset (recommended)
4. Organize multiple city datasets (small / medium / large)
5. Use the datasets with `trafficCAR`

The recommended workflow is:

> **GeoJSON (raw, portable) → `.rda` (fast, CRAN-friendly)**

---

## Step 1: Obtain road data for a city

A common source of road data is **OpenStreetMap**, accessed via the `osmdata`
package.

```{r}
library(osmdata)
library(sf)
```

Choose a city name or bounding box.

```{r, eval = FALSE}
city <- "College Station, Texas, USA"

q <- opq(city) |>
  add_osm_feature(key = "highway")

osm <- osmdata_sf(q)
```

Extract road geometries:

```{r, eval = FALSE}
roads <- osm$osm_lines
```

At this point, `roads` is an `sf` object with LINESTRING/MULTILINESTRING
geometries.

---

## Step 2: Clean and inspect geometries

It is recommended to keep only a small number of attributes.

```{r, eval = FALSE}
roads <- roads[, c("highway", "name", "geometry")]
unique(st_geometry_type(roads))
```

Only `LINESTRING` and `MULTILINESTRING` geometries are supported by
`trafficCAR`.

---

## Step 3: Write the dataset to GeoJSON

Save the road network to a GeoJSON file.

```{r, eval = FALSE}
st_write(
  roads,
  "roads_cstat.geojson",
  delete_dsn = TRUE,
  quiet = TRUE
)
```

This file is:
- human-readable
- portable outside R
- suitable as a long-term raw data source

---

## Step 4: Convert GeoJSON to `.rda` (recommended)

For regular use with `trafficCAR`, convert GeoJSON to `.rda`.
This provides faster loading and avoids file I/O in examples.

```{r, eval = FALSE}
library(usethis)

roads_cstat <- st_read("roads_cstat.geojson", quiet = TRUE)
use_data(roads_cstat, overwrite = TRUE)
```

This creates:

```
data/roads_cstat.rda
```

You can now load the dataset with:

```{r, eval = FALSE}
data(roads_cstat)
```

---

## Step 5: Use the dataset in trafficCAR

```{r, eval = FALSE}
library(trafficCAR)

segs <- roads_to_segments(
  roads_cstat,
  split_at_intersections = TRUE
)

adj <- build_adjacency(segs)
```

The resulting segments and adjacency matrix can be used directly in
model fitting.

---

## Organizing multiple city datasets (small / medium / large)

When working with multiple cities or scales, a consistent naming convention
is strongly recommended.

### Suggested naming scheme

Use both **city** and **scale** in the object name:

```
roads_<city>_<scale>
```

Examples:
- `roads_cstat_small`
- `roads_cstat_medium`
- `roads_cstat_large`
- `roads_chicago_small`

### Example directory layout

```
inst/extdata/
  roads_cstat_small.geojson
  roads_cstat_medium.geojson
  roads_cstat_large.geojson

data-raw/
  roads_cstat_small.R
  roads_cstat_medium.R
  roads_cstat_large.R

data/
  roads_cstat_small.rda
  roads_cstat_medium.rda
  roads_cstat_large.rda
```

Each `data-raw/*.R` file should contain exactly one dataset and one call to
`use_data()`.

### Example `data-raw` script

```{r eval=FALSE}
library(sf)
library(usethis)

roads_cstat_small <- st_read(
  "inst/extdata/roads_cstat_small.geojson",
  quiet = TRUE
)

use_data(roads_cstat_small, overwrite = TRUE)
```

---

## Using external GeoJSON files directly

The following example shows how a **user project** (outside the package)
may pass a GeoJSON file path directly. This code is shown for illustration
only and is **not run** when building the vignette.

```{r eval=FALSE}
segs <- roads_to_segments("roads_cstat.geojson")
```

Within the package itself (examples and vignettes), `.rda` datasets are used
to avoid file I/O and ensure CRAN compatibility.

---

## Summary

- Use **GeoJSON** as the authoritative raw format
- Convert to **`.rda`** for routine use with `trafficCAR`
- Adopt a clear naming scheme for multiple cities and scales
- Load `.rda` datasets in examples and vignettes

This workflow keeps projects reproducible, scalable, and CRAN-friendly.
