---
title: "trafficCAR-intro"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{trafficCAR-intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`trafficCAR` provides tools for constructing **conditional autoregressive (CAR)**
precision matrices on graphs derived from road networks. The package supports
network creation from spatial linework, basic spatial weight construction, and
proper/ICAR precision matrices that can be used in Gaussian hierarchical
models.

This vignette walks through a minimal workflow that uses the bundled example
roads data to build a network, create spatial weights, and construct CAR/ICAR
precision matrices suitable for simulation or modeling.

## Load the package and example data

The package ships with a small GeoJSON file representing a handful of road
segments. We will read it with **sf** and feed it into the network builder.

```{r}
library(trafficCAR)
library(Matrix)

road_file <- system.file("extdata", "roads.geojson", package = "trafficCAR")
if (!nzchar(road_file) || !file.exists(road_file)) {
  stop("Example road data not found in inst/extdata.")
}

roads_sf <- sf::st_read(road_file, quiet = TRUE)
```

## Build a road network graph

`build_network()` cleans the linework, projects it to a metric coordinate
reference system, and returns a list containing nodes, edges, an igraph object,
and a sparse adjacency matrix.

```{r}
net <- build_network(
  roads_sf,
  crs_out = 3857,
  node_intersections = TRUE,
  snap_tol = 0,
  simplify = TRUE
)

c(
  nodes = nrow(net$nodes),
  edges = nrow(net$edges),
  adjacency_nnz = Matrix::nnzero(net$A)
)
```

The adjacency matrix encodes undirected connectivity between road nodes (one per
unique endpoint). Self-loops and duplicate edges are removed when
`simplify = TRUE`.

## Spatial weights from the adjacency matrix

You can work directly with the binary adjacency matrix or derive alternative
weights. Row-standardizing the adjacency produces a matrix whose rows sum to 1
for non-isolated nodes.

```{r}
W_row <- weights_from_adjacency(net$A, style = "row-standardized")

# first few row sums (isolated nodes, if any, remain at 0)
Matrix::rowSums(W_row)[1:6]
```

For CAR precision matrices we will use the binary adjacency `net$A` to keep the
interpretation simple.

## CAR and ICAR precision matrices

With an adjacency in hand, CAR precision matrices follow the common form
\(Q = \tau (D - \rho A)\), where \(D\) is the degree matrix. The helper
`car_precision()` builds both proper and intrinsic variants.

```{r}
# Proper CAR with moderate spatial dependence
Q_car <- car_precision(net$A, type = "proper", rho = 0.25, tau = 1)

# Intrinsic CAR (ICAR) with optional Besag scaling
Q_icar <- intrinsic_car_precision(net$A, tau = 1, scale = TRUE)

Q_car[1:6, 1:6]
Matrix::rankMatrix(Q_icar)
```

`Q_car` is sparse and positive definite for \(|\rho| < 1\) on graphs without
isolated nodes. The ICAR precision `Q_icar` is singular with rank deficiency
equal to the number of connected components; Besag scaling rescales it so that
the geometric mean marginal variance is one.

## Sampling from a CAR precision matrix

`trafficCAR` includes an internal helper for drawing Gaussian random effects
without forming a dense covariance matrix. This is useful for simulation studies
or for checking that a chosen \(\rho\) produces sensible variability.

```{r}
set.seed(123)

theta <- trafficCAR:::rmvnorm_prec(Q_car)
c(mean = mean(theta), sd = sd(theta))
```

For ICAR models, a small ridge can be added before sampling to avoid numerical
issues, followed by centering to enforce identifiability. This mirrors the
approach shown in the sampling vignette.

```{r}
if (nrow(Q_icar) > 0) {
  eps <- 1e-6
  Q_icar_ridge <- Q_icar + Matrix::Diagonal(nrow(Q_icar), x = eps)

  theta_icar <- trafficCAR:::rmvnorm_prec(Q_icar_ridge)
  theta_icar <- theta_icar - mean(theta_icar)
  summary(theta_icar)
}
```

## Interactive road visualizations

`trafficCAR` includes lightweight interactive mapping helpers built on
**leaflet**. They visualize segment-level quantities (e.g., posterior means or
predicted traffic levels) on the bundled example road network.

The example below fabricates a simple score for each segment so the map can be
rendered without running a full model. When knitting with the optional
dependencies installed (`leaflet` and `viridisLite`), `map_roads_interactive()`
produces an interactive widget with tooltips and a continuous legend.

```{r}
road_file <- system.file("extdata", "roads.geojson", package = "trafficCAR")
roads_sf <- sf::st_read(road_file, quiet = TRUE)

# mock segment-level values (replace with posterior summaries in real workflows)
set.seed(123)
roads_sf$traffic_score <- runif(nrow(roads_sf), min = 0, max = 1)

has_leaflet <- requireNamespace("leaflet", quietly = TRUE) &&
  requireNamespace("viridisLite", quietly = TRUE)

if (has_leaflet) {
  map_roads_interactive(roads_sf, value_col = "traffic_score")
} else {
  message("Install 'leaflet' and 'viridisLite' to view the interactive map.")
}
```

You can also expose multiple layers (e.g., mean and interval width) using
`map_roads_interactive_layers()`. Here we create a synthetic interval width and
enable a layers control so readers can toggle between the two overlays.

```{r}
# mock interval width derived from the fabricated score
roads_sf$interval_width <- pmax(0, roads_sf$traffic_score - 0.5)

if (has_leaflet) {
  map_roads_interactive_layers(
    roads_sf,
    layers = c("Score" = "traffic_score", "Interval width" = "interval_width")
  )
}

```

## Takeaways

- `build_network()` converts road linework to a graph with adjacency and edge
  metadata.
- `weights_from_adjacency()` can produce binary or row-standardized weights.
- `car_precision()` and `intrinsic_car_precision()` create sparse CAR/ICAR
  precision matrices ready for modeling.
- The internal sampler `rmvnorm_prec()` enables quick Gaussian draws from those
  precision matrices without forming covariances.

## Further directions

Planned extensions include:

- Poisson and negative binomial models for traffic volume
- Layered interactive maps (raw vs modeled traffic)
- Shiny applications for time-varying traffic data
